---
title: "FCS Clean-up summary"
output:
  html_document:
    df_print: paged
---

```{r, include=FALSE}
library(tidyverse)

knitr::opts_chunk$set(fig.width=13, fig.height=7, fig_caption=TRUE)

ps <- snakemake@params

MIN_EVENT_SOP1 <- ps[["min_events"]][["sop1"]]
MIN_EVENT_SOP2 <- ps[["min_events"]][["sop2"]]
MIN_EVENT_SOP3 <- ps[["min_events"]][["sop3"]]
VDIFF_LIMIT <- ps[["detector_limits"]][["voltage"]]
GDIFF_LIMIT <- ps[["detector_limits"]][["gain"]]
TRUNCATION <- as.double(ps[["edge_thresh"]])

df_meta_issues <- read_tsv(
  snakemake@input[["issues"]],
  col_types = cols(
    file_index = "i",
    sop = "i",
    eid = "i",
    om = "c",
    group = "c",
    missing_file = "l",
    has_voltage_variation = "l",
    has_gain_variation = "l",
    has_insufficient_events = "l",
    has_multi_serial = "l",
    has_multi_cytometer = "l",
    has_multi_system = "l",
    missing_time = "l",
    missing_colors = "l",
    missing_scatter_area = "l",
    missing_scatter_height = "l",
    has_incomplete_set = "l",
    has_manual_exclusion = "l",
    .default = "-"
  )
) %>%
  mutate(
    group = fct_relevel(
      group,
      "SOP 3: Test Pheno",
      "SOP 3: Test Count",
      "SOP 3: Test FMO",
      "SOP 3: QC Pheno",
      "SOP 3: QC Count",
      after = 10
    )
  ) %>%
  mutate(om = fct_rev(om))

df_voltgain <- read_tsv(
  snakemake@input[["voltgain"]],
  col_types = "icccccccdd"
) %>%
  mutate(om = fct_rev(om))

df_ranges <- read_tsv(
  snakemake@input[["edge_removed"]],
  col_types = "icid",
  col_names = c("file_index", "channel", "nover", "fracover")
) %>%
  left_join(df_meta_issues, by = "file_index")

df_all_issues <- df_meta_issues %>%
  mutate(
    fatal_metadata_error = case_when(
      missing_file ~ "1 missing file",
      has_insufficient_events ~ "2 too few events",
      has_manual_exclusion ~ "3 manual exclusion",
      missing_time ~ "4 missing time channel" ,
      missing_scatter_area ~ "5 missing scatter area",
      missing_scatter_height ~ "6 missing scatter height",
      missing_colors ~ "7 missing color channel",
      has_incomplete_set ~ "8 incomplete",
      TRUE ~ "9 no fatal metadata error"
    ) %>%
      fct_relabel(~ str_sub(.x, 3)),
    nonfatal_metadata_error = case_when(
      has_multi_serial | has_multi_cytometer | has_multi_system ~ "1 multi system",
      fatal_metadata_error != "no fatal metadata error" ~ "2 previous error",
      TRUE ~ "4 no non-fatal metadata error"
    ) %>%
      fct_relabel(~ str_sub(.x, 3)),
    channel_error = case_when(
      missing_time ~ "1 missing time channel" ,
      missing_scatter_area ~ "2 missing scatter area",
      missing_scatter_height ~ "3 missing scatter height",
      missing_colors ~ "4 missing color channel",
      fatal_metadata_error != "no fatal metadata error" ~ "5 previous error",
      TRUE ~ "6 no channel error"
    ) %>%
      fct_relabel(~ str_sub(.x, 3)),
    fatal_metadata_alpha = if_else(fatal_metadata_error == "no fatal metadata error", 0.5, 1),
    nonfatal_metadata_alpha = if_else(nonfatal_metadata_error == "no non-fatal metadata error", 0.5, 1),
    channel_alpha = if_else(channel_error == "no channel error", 0.5, 1),
  )
```

This is a summary of the errors found during FCS file cleanup.

The minimum events for each SOP were deemed as follows:

* SOP 1: `r as.integer(MIN_EVENT_SOP1)`
* SOP 2: `r as.integer(MIN_EVENT_SOP2)`
* SOP 3: `r as.integer(MIN_EVENT_SOP3)`

# Voltage/Gain variation

All voltages for each channel for each machine should be the same compared to
SOP-1.

```{r, echo = FALSE, results = "asis"}
df_voltgain %>%
  ggplot(aes(vdiff, om, color = std_name_long)) +
  geom_jitter(width = 0.005, height = 0.1, size = 0.5, na.rm= TRUE) +
  facet_wrap(c("group"), nrow = 1) +
  labs(x = "Relative $PnV difference from SOP-1 ((x - x1) / x1)", y = NULL, color = "Channel")

df_voltgain %>%
  ggplot(aes(gdiff, om, color = std_name_long)) +
  geom_jitter(width = 0, height = 0.1, size = 0.5, na.rm= TRUE) +
  facet_wrap(c("group"), nrow = 1) +
  labs(x = "Relative $PnG difference from SOP-1 ((x - x1) / x1)", y = NULL, color = "Channel")
```

# Fatal Metadata Errors

Error explanation:

1. missing file: FCS file is completely missing from dataset
2. too few events: FCS file has less than the minimum required events as
   dictated by the SOP (before filtering as done with the time channel below)
3. manual exclusion: file was excluded manually for a miscellaneous reason
4. missing time channel: time channel is absent from the FCS file
5. missing scatter area: missing either FSC-A or SSC-A
6. missing scatter height: missing either FSC-H or SSC-H (cell samples only)
7. missing color channels: at least 1 of the 8 standardized color channels is
   missing
8. incomplete: FCS file is in a dataset that is incomplete, which implies
   another file in the dataset has one of the below errors
   
For 8, "incomplete" datasets were determined as those with at least one file
missing in SOP 2 - Matrix 2 or SOP 3 Test Phenotype, since without these it is
impossible to carry out the objective of the analysis.

```{r, echo = FALSE, results = "asis"}
df_all_issues %>%
  ggplot(aes(y = om, fill = fatal_metadata_error, alpha = fatal_metadata_alpha)) +
  geom_bar() +
  facet_wrap(c("group"), nrow = 1) +
  scale_alpha_continuous(limits = c(0, 1)) +
  labs(x = "Number of Files", y = NULL, fill = "Fatal\nMetadata Error") +
  theme(legend.position = "bottom") +
  guides(alpha = "none")
```

# Non-Fatal Metadata Errors

Error explanation:

1. multi system: dataset may have been acquired on multiple machines (as shown
   via $CYT, $CYTSN, or $SYS keywords)
2. previous error: a fatal error from the above section
   
```{r, echo = FALSE, results = "asis"}
df_all_issues %>%
  ggplot(aes(y = om, fill = nonfatal_metadata_error, alpha = nonfatal_metadata_alpha)) +
  geom_bar() +
  facet_wrap(c("group"), nrow = 1) +
  scale_alpha_continuous(limits = c(0, 1)) +
  labs(x = "Number of Files", y = NULL, fill = "Non-Fatal\nMetadata Error") +
  theme(legend.position = "bottom") +
  guides(alpha = "none")
```

# Truncation/range summary

Some channels contain events that are at/near their maximum range.

Events were considered truncated if they were greater than `r (TRUNCATION * 100)`
percent of the range given by $PnR for the given channel.

Note that only files that contained no fatel or channel errors were processed
at this step.

## Scatter

```{r, echo = FALSE, results = "asis"}
df_ranges %>%
  filter(channel %in% c("fsc_a", "fsc_h", "ssc_a", "ssc_h")) %>%
  ggplot(aes(fracover, om, color = channel)) +
  geom_jitter(width = 0, height = 0.1, size = 0.5) +
  facet_wrap(c("group"), nrow = 1) +
  labs(x = "Truncation Rate", y = NULL, fill = "Channel") +
  theme(legend.position = "bottom") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_x_continuous(limit = c(0, 1), labels = scales::percent, breaks = c(0, 0.5, 1))
```

## Color

```{r, echo = FALSE, results = "asis"}
df_ranges %>%
  filter(!channel %in% c("fsc_a", "fsc_h", "ssc_a", "ssc_h")) %>%
  ggplot(aes(fracover, om, color = channel)) +
  geom_jitter(width = 0, height = 0.1, size = 0.5) +
  facet_wrap(c("group"), nrow = 1) +
  labs(x = "Truncation Rate", y = NULL, fill = "Channel") +
  theme(legend.position = "bottom") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_x_continuous(limit = c(0, 1), labels = scales::percent, breaks = c(0, 0.5, 1))
```
