---
title: "FCS Clean-up summary"
output:
  html_document:
    df_print: paged
---

```{r, include=FALSE}
library(tidyverse)

knitr::opts_chunk$set(fig.width=13, fig.height=7, fig_caption=TRUE)

ps <- snakemake@params

MIN_EVENT_SOP1 <- ps[["min_events"]][["sop1"]]
MIN_EVENT_SOP2 <- ps[["min_events"]][["sop2"]]
MIN_EVENT_SOP3 <- ps[["min_events"]][["sop3"]]
VDIFF_LIMIT <- ps[["detector_limits"]][["voltage"]]
GDIFF_LIMIT <- ps[["detector_limits"]][["gain"]]
TRUNCATION <- as.double(ps[["edge_thresh"]])

df_meta_issues <- read_tsv(
  snakemake@input[["issues"]],
  col_types = cols(
    file_index = "i",
    sop = "i",
    eid = "i",
    om = "c",
    group = "c",
    missing_file = "l",
    has_voltage_variation = "l",
    has_gain_variation = "l",
    has_insufficient_events = "l",
    has_multi_serial = "l",
    has_multi_cytometer = "l",
    has_multi_system = "l",
    missing_time = "l",
    missing_colors = "l",
    missing_scatter_area = "l",
    missing_scatter_height = "l",
    has_incomplete_set = "l",
    has_manual_exclusion = "l",
    .default = "-"
  )
)

df_voltgain <- read_tsv(
  snakemake@input[["voltgain"]],
  col_types = "icccccccdd"
)

df_ranges <- read_tsv(
  snakemake@input[["edge_removed"]],
  col_types = "icid",
  col_names = c("file_index", "channel", "nover", "fracover")
) %>%
  left_join(df_meta_issues, by = "file_index")

df_all_issues <- df_meta_issues %>%
  mutate(
    fatal_metadata_error = case_when(
      missing_file ~ "1 missing file",
      has_insufficient_events ~ "2 too few events",
      has_manual_exclusion ~ "3 manual exclusion",
      has_incomplete_set ~ "4 incomplete",
      TRUE ~ "5 no fatal metadata error"
    ) %>%
      fct_relabel(~ str_sub(.x, 3)),
    nonfatal_metadata_error = case_when(
      has_multi_serial | has_multi_cytometer | has_multi_system ~ "1 multi system",
      fatal_metadata_error != "no fatal metadata error" ~ "2 previous error",
      TRUE ~ "4 no non-fatal metadata error"
    ) %>%
      fct_relabel(~ str_sub(.x, 3)),
    channel_error = case_when(
      missing_time ~ "1 missing time channel" ,
      missing_scatter_area ~ "2 missing scatter area",
      missing_colors ~ "3 missing color channel",
      fatal_metadata_error != "no fatal metadata error" ~ "4 previous error",
      TRUE ~ "5 no channel error"
    ) %>%
      fct_relabel(~ str_sub(.x, 3)),
    fatal_metadata_alpha = if_else(fatal_metadata_error == "no fatal metadata error", 0.5, 1),
    nonfatal_metadata_alpha = if_else(nonfatal_metadata_error == "no non-fatal metadata error", 0.5, 1),
    channel_alpha = if_else(channel_error == "no channel error", 0.5, 1),
  )
```

This is a summary of the errors found during FCS file cleanup.

The minimum events for each SOP were deemed as follows:

* SOP 1: `r as.integer(MIN_EVENT_SOP1)`
* SOP 2: `r as.integer(MIN_EVENT_SOP2)`
* SOP 3: `r as.integer(MIN_EVENT_SOP3)`

# Voltage/Gain variation

All voltages for each channel for each machine should be the same compared to
SOP-1.

```{r, echo = FALSE, results = "asis"}
df_voltgain %>%
  ggplot(aes(vdiff, om, color = std_name_long)) +
  geom_jitter(width = 0.005, height = 0.1, size = 0.5, na.rm= TRUE) +
  facet_wrap(c("group"), nrow = 1) +
  labs(x = "Relative $PnV difference from SOP-1 ((x - x1) / x1)", y = NULL, color = "Channel")

df_voltgain %>%
  ggplot(aes(gdiff, om, color = std_name_long)) +
  geom_jitter(width = 0, height = 0.1, size = 0.5, na.rm= TRUE) +
  facet_wrap(c("group"), nrow = 1) +
  labs(x = "Relative $PnG difference from SOP-1 ((x - x1) / x1)", y = NULL, color = "Channel")
```

# Fatal Metadata Errors

Error explanation:

1. missing file: FCS file is completely missing from dataset
2. too few events: FCS file has less than the minimum required events as
   dictated by the SOP (before filtering as done with the time channel below)
3. voltage/gain variation: voltage and/or gain within each channel has
   unacceptable variation
4. incomplete: FCS file is in a dataset that is incomplete, which implies
   another file in the dataset has one of the below errors
   
For 3, variation was computed as the relative difference of each non-SOP-1 FCS
file to those in SOP-1 (ie (x - x1) / x1). The unacceptable limits for voltage
and gain were `r VDIFF_LIMIT` and `r GDIFF_LIMIT` respectively.

For 4, "incomplete" datasets were determined as those with at least one file
missing in SOP 2 - Matrix 2 or SOP 3 Test (Phenotype or FMO), since without
these it is impossible to carry out the objective of the analysis.

```{r, echo = FALSE, results = "asis"}
df_all_issues %>%
  ggplot(aes(y = om, fill = fatal_metadata_error, alpha = fatal_metadata_alpha)) +
  geom_bar() +
  facet_wrap(c("group"), nrow = 1) +
  scale_alpha_continuous(limits = c(0, 1)) +
  labs(x = "Number of Files", y = NULL, fill = "Fatal\nMetadata Error") +
  theme(legend.position = "bottom") +
  guides(alpha = "none")
```

# Non-Fatal Metadata Errors

Error explanation:

1. multi system: dataset may have been acquired on multiple machines (as shown
   via $CYT, $CYTSN, or $SYS keywords)
2. previous error: a fatal error from the above section
   
```{r, echo = FALSE, results = "asis"}
df_all_issues %>%
  ggplot(aes(y = om, fill = nonfatal_metadata_error, alpha = nonfatal_metadata_alpha)) +
  geom_bar() +
  facet_wrap(c("group"), nrow = 1) +
  scale_alpha_continuous(limits = c(0, 1)) +
  labs(x = "Number of Files", y = NULL, fill = "Non-Fatal\nMetadata Error") +
  theme(legend.position = "bottom") +
  guides(alpha = "none")
```

# Channel Errors

Error explanation:

1. missing time channel: time channel is absent from the FCS file
2. missing scatter area: missing either FSC-A or SSC-A
3. missing scatter height: missing either FSC-H or SSC-H
4. missing color channels: at least 1 of the 8 standardized color channels is
   missing
5. previous error: a fatal error from the above section
   
```{r, echo = FALSE, results = "asis"}
df_all_issues %>%
  ggplot(aes(y = om, fill = channel_error, alpha = channel_alpha)) +
  geom_bar() +
  facet_wrap(c("group"), nrow = 1) +
  scale_alpha_continuous(limits = c(0, 1)) +
  labs(x = "Number of Files", y = NULL, fill = "Channel Error") +
  theme(legend.position = "bottom") +
  guides(alpha = "none")
```

# Truncation/range summary

Some channels contain events that are at/near their maximum range.

Events were considered truncated if they were greater than `r (TRUNCATION * 100)`
percent of the range given by $PnR for the given channel.

Note that only files that contained no fatel or channel errors were processed
at this step.

## Scatter

```{r, echo = FALSE, results = "asis"}
df_ranges %>%
  filter(channel %in% c("fsc_a", "fsc_h", "ssc_a", "ssc_h")) %>%
  ggplot(aes(fracover, om, color = channel)) +
  geom_jitter(width = 0, height = 0.1, size = 0.5) +
  facet_wrap(c("group"), nrow = 1) +
  labs(x = "Truncation Rate", y = NULL, fill = "Channel") +
  theme(legend.position = "bottom") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_x_continuous(limit = c(0, 1), labels = scales::percent, breaks = c(0, 0.5, 1))
```

## Color

```{r, echo = FALSE, results = "asis"}
df_ranges %>%
  filter(!channel %in% c("fsc_a", "fsc_h", "ssc_a", "ssc_h")) %>%
  ggplot(aes(fracover, om, color = channel)) +
  geom_jitter(width = 0, height = 0.1, size = 0.5) +
  facet_wrap(c("group"), nrow = 1) +
  labs(x = "Truncation Rate", y = NULL, fill = "Channel") +
  theme(legend.position = "bottom") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_x_continuous(limit = c(0, 1), labels = scales::percent, breaks = c(0, 0.5, 1))
```
