---
title: "FCS files with time issues"
output:
  html_document:
    df_print: paged
---

```{r, include=FALSE}
library(tidyverse)
library(scattermore)

knitr::opts_chunk$set(fig.width=13, fig.height=7, fig_caption=TRUE)

## process_gates <- function(df, gt) {
##   df %>%
##     group_by(file_index) %>%
##     mutate(novalid = sum(valid) == 0) %>%
##     mutate(hasgates = n() > 1) %>%
##     ungroup() %>%
##     add_column(gate_type = gt)
## }

df_gates <- read_tsv(
  snakemake@input[["flat"]],
  col_types = "iii--lil",
  col_names = c("file_index", "start", "end", "valid", "parent", "is_break")
) %>%
  group_by(file_index) %>%
  mutate(novalid = sum(valid) == 0) %>%
  mutate(hasgates = n() > 1) %>%
  ungroup() %>%
  mutate(gate_type = factor(if_else(is_break, "flow rate", "anomaly")))
  

# ASSUME these are sorted by file/event index
df_events <- read_tsv(
  snakemake@input[["events"]],
  col_types = "iid",
  col_names = c("file_index", "event_index", "time")
)

df_meta <- read_tsv(
  snakemake@input[["meta"]],
  col_types = cols(
    machine = "c",
    org = "c",
    file_index = "i",
    material = "c",
    sop = "i",
    eid = "i",
    rep = "i",
    .default = "-"
  )
)

df_meta_issues <- read_tsv(
  snakemake@input[["issues"]],
  col_types = cols(
    file_index = "i",
    sop = "i",
    eid = "i",
    om = "c",
    group = "c",
    missing_file = "l",
    has_voltage_variation = "l",
    has_gain_variation = "l",
    has_insufficient_events = "l",
    missing_time = "l",
    missing_colors = "l",
    missing_scatter_area = "l",
    missing_scatter_height = "l",
    has_incomplete_set = "l",
    .default = "-"
  )
) %>%
  mutate(
    group = fct_relevel(
      group,
      "SOP 3: Test Pheno",
      "SOP 3: Test Count",
      "SOP 3: Test FMO",
      "SOP 3: QC Pheno",
      "SOP 3: QC Count",
      after = 10
    )
  ) %>%
  mutate(om = fct_rev(om))

make_plot <- function(df_gates, key, both) {
  i <- key$file_index[[1]]
  m <- df_meta %>%
    filter(file_index == i) %>%
    head(n = 1)
  cat(sprintf("### %s - %s\n\n", m$org, m$machine))
  cat(sprintf("File Index = %d\n\n", i))
  cat(sprintf("SOP = %d\n\n", m$sop))
  cat(sprintf("Experiment = %d\n\n", m$eid))
  cat(sprintf("Rep = %d\n\n", m$rep))
  cat(sprintf("Material = %s\n\n", m$material))
  p <- df_events %>%
    filter(file_index == i) %>%
    ggplot(aes(event_index, time)) +
    geom_scattermore(pointsize=2)
  # TODO add colors for each gate type
  if (FALSE) {
    rest <- list(
      geom_rect(
        data = df_gates,
        aes(xmin = start, xmax = end, ymin = 0, ymax = Inf, fill = gate_type),
        alpha = 0.2,
        inherit.aes = FALSE
      ),
      scale_fill_discrete(drop = FALSE),
      labs(x = "Event", y = "Time Channel", fill = "Gate Type")
    )
  } else {
    rest <- list(
      geom_vline(
        data = slice_tail(df_gates, n = -1),
        aes(xintercept = start),
        color = "red",
        linetype = "dashed"
      ),
      labs(x = "Event", y = "Time Channel")
    )
  }
  print(p + rest)
  cat("\n\n")
}

anomaly_idx <- df_gates %>%
  filter(!is_break) %>%
  filter(parent > 0) %>%
  pull(file_index)

bad_time_idx <- unique(df_events$file_index)

df_time_issues <- df_meta_issues %>%
  mutate(
    has_time_anomaly = file_index %in% bad_time_idx & !(file_index %in% anomaly_idx),
    has_time_nonlinear = file_index %in% bad_time_idx & file_index %in% anomaly_idx,
    time_error = case_when(
      missing_file | has_voltage_variation | has_gain_variation |
        has_insufficient_events | missing_time | missing_colors |
        missing_scatter_area | missing_scatter_height | has_incomplete_set ~
          "1 previous error",
      has_time_anomaly ~ "2 has time anomaly",
      has_time_nonlinear ~ "3 has uneven flow rate",
      TRUE ~ "4 no time error"
    ) %>%
      fct_relabel(~ str_sub(.x, 3)),
    time_alpha = if_else(time_error == "no time error", 0.5, 1)
  )

```

The following are the time channels of the FCS files with detected issues 
(further elaborated below).

# Error Summary

```{r, echo = FALSE, results = "asis"}
df_time_issues %>%
  ggplot(aes(y = om, fill = time_error, alpha = time_alpha)) +
  geom_bar() +
  facet_wrap(c("group"), nrow = 1) +
  scale_alpha_continuous(limits = c(0, 1)) +
  labs(x = "Number of Files", y = NULL, fill = "Time Error") +
  theme(legend.position = "bottom") +
  guides(alpha = "none")
```


# Invalid Files

In each of these cases, the minimum nunmber of required events was not
achieved after trying to gate around these issues.

## Files with anomalies

These files have one or both of these issues:

* (large) non-monotonicity: the time channel reset to (near) zero in the middle
  of the run

* gaps: the time channel has a large pause in the middle, which usually
  indicates a clog or bubble

```{r, echo = FALSE, results = "asis"}
.df_ano <- df_gates %>%
  filter(novalid) %>%
  filter(!is_break)

if (nrow(.df_ano) > 0) {
  .df_ano %>%
    group_by(file_index) %>%
    group_walk(~ make_plot(.x, .y, FALSE))
} else {
  cat("None\n\n")
}

```

## Files with non-linearity

These files have large deviations in flow rate somewhere in the middle.

```{r, echo = FALSE, results = "asis"}
.df_flat <- df_gates %>%
  filter(novalid) %>%
  filter(is_break)

if (nrow(.df_flat) > 0) {
  .df_flat %>%
    group_by(file_index) %>%
    group_walk(~ make_plot(.x, .y, FALSE))
} else {
  cat("None\n\n")
}
```

# Valid Files

These are files that have at least one gate but exceeded the minimum event
count in at least one gate.

```{r, echo = FALSE, results = "asis"}
df_gates %>%
  filter(!novalid) %>%
  filter(hasgates) %>%
  arrange(file_index) %>%
  group_by(file_index) %>%
  group_walk(~ make_plot(.x, .y, TRUE))
```
