---
title: "FCS files with time issues"
output:
  html_document:
    df_print: paged
---

```{r, include=FALSE}
library(tidyverse)
library(scattermore)

knitr::opts_chunk$set(fig.width=8, fig.height=4, fig_caption=TRUE)

anomaly_indices <- unique(read_tsv(snakemake@input[["anomaly"]], col_types = "i---")[[1]])

# ASSUME these are sorted by file/event index
df_events <- read_tsv(
  snakemake@input[["events"]],
  col_types = "iid",
  col_names = c("file_index", "event_index", "time")
) %>%
  mutate(passed_anomaly = file_index %in% anomaly_indices)

df_meta <- read_tsv(
  snakemake@input[["meta"]],
  col_types = cols(
    machine = "c",
    org = "c",
    file_index = "i",
    material = "c",
    sop = "i",
    exp = "i",
    rep = "i",
    .default = "-"
  )
)

make_plot <- function(df_events, key) {
  i <- key$file_index
  m <- df_meta %>%
    filter(file_index == i) %>%
    head(n = 1)
  cat(sprintf("### %s - %s\n\n", m$org, m$machine))
  cat(sprintf("File Index = %d\n\n", i))
  cat(sprintf("SOP = %d\n\n", m$sop))
  cat(sprintf("Experiment = %d\n\n", m$exp))
  cat(sprintf("Rep = %d\n\n", m$rep))
  cat(sprintf("Material = %s\n\n", m$material))
  p <- df_events %>%
    ggplot(aes(event_index, time)) +
    geom_scattermore(pointsize=2) +
    labs(x = "Event", y = "Time Channel")
  print(p)
  cat("\n\n")
}

```

The following are the time channels of the FCS files with detected issues 
(further elaborated below).

In each of these cases, the minimum nunmber of required events was not
achieved after trying to gate around these issues.

# Files with anomalies

These files have one or both of these issues:

* (large) non-monotonicity: the time channel reset to (near) zero in the middle
  of the run

* gaps: the time channel has a large pause in the middle, which usually
  indicates a clog or bubble

```{r, echo = FALSE, results = "asis"}
df_events %>%
  filter(!passed_anomaly) %>%
  group_by(file_index) %>%
  group_walk(make_plot)
```

# Files with non-linearity

These files have large deviations in flow rate somewhere in the middle.

```{r, echo = FALSE, results = "asis"}
df_events %>%
  filter(passed_anomaly) %>%
  group_by(file_index) %>%
  group_walk(make_plot)
```
